Safety Cases for Software Product Lines: Terminology and Example

### 2017-10-19 doc/SafetyCases/SafetyCasesPL.tex
minor document updates

Several attempts have been made to leverage the high degree of commonality across products of a Software Product Line (SPL) to minimize the cost of product line analysis in general, and safety analysis in particular. This document explores previous attempts and fundamental questions at the crossroads of Safety Analysis and Software Product Lines.

\subsection{Goals}
This document aims at:
\begin{itemize}
\item Introducing terminology for both Software Product Line and Safety Case concepts
\item Demonstrating the Product Line and Safety Case concepts on a simple exemplar
\item To try to come to a mutual agreement on concepts, terminology and the Product Line safety engineering at GM
\end{itemize}

\subsection{Non-Goals}

@@

Throughout this document we use a Fuel Level Display Product Line (FLDPL) example from \cite{Gallucci_2013}. This is a simply product line safety case that extends a single product ISO26262 safety case \cite{Dardar_2014}. The diagrams in this document also come from \cite{Gallucci_2013} unless otherwise specified.

@@

We start with some basic definitions, together with some simple examples illustrating them. The definitions and the object store example presented here are adapted from those in \cite{Thum}.

@@

\item[Software Product Line]
A Software Product Line (SPL) is a set of similar software products built from a common set of artifacts (e.g. design/specification models, code constructs, test cases, etc...).

@@

\item[Feature]
A feature is an externally visible property, aspect or quality of a software system.

@@

\item[Feature Diagram]
a hierarchical graphical representation of a Feature Model, where each non-root feature depends on its parent.

@@

For example, figure \ref{fig:FeatureDiagram} shows a feature diagram of the FLDPL. In this diagram, the usage context is composed of two mandatory features: Vehicle type and Fuel type. Vehicle type can be either Truck or Bus (mutually exclusive options). Similarly, Fuel type can be either Liquid or Gas. In addition, the FLD system has one mandatory feature (Fuel Level Estimation) and one optional feature (Low Fuel Level Warning). An annotation (O1) is added to the diagram to specify the set of products where the low fuel level warning feature is present.

@@

\item[Presence Condition]
A Presence Condition (PC) is a boolean formula over the set of features that represents the set of products in which an artifact exists.
\end{description}

@@

\item[Product Configuration]
A product configuration is a boolean formula over the set of features of an SPL that specifies the set of features present in a product.
\end{description}

@@

\item[150\% Representation]
The 150\% representation of an SPL is a syntactic representation with syntactic constructs belonging to all the features of the SPL. 
\end{description}

Since in the vast majority of cases all those syntactic constructs can not all coexist in a product (given feature model constraints), this representation is referred to as 150\%, as opposed to the 100\% representation of a single product.

Examples of 150\% representations can be found in section 4.

@@

\lstset{
	basicstyle=\ttfamily,
     keywordstyle=\color{blue}\ttfamily,
     stringstyle=\color{red}\ttfamily,
     commentstyle=\color{green}\ttfamily,
     morecomment=[l][\color{magenta}]{\#},
	numbers=left, 
	tabsize=2,
	captionpos=b, 
	frame=shadowbox
}

@@

\begin{lstlisting}[language=Java, caption=Generative FLDPL, label={lst:Generative}, float]
// Feature module FLDGas
class FLD {
	void reportFuelLevel(float level) { 
		dashboard.display(level); 
	}
}

@@

@@

\begin{lstlisting}[language=Java, caption=Generated FLD Product, label={lst:Generated}, float]
class FLD {
	const float THRESHOLD;

Given the product configuration (Liquid $\wedge$ Truck $\wedge$ FuelWarning), the product in listing \ref{lst:Generated} is generated.

@@

If we are to rewrite FLDPL using CPP macros, it would look like listing \ref{lst:Annotative}. A single feature implementation (e.g. LOW\_GAS\_WARNING is now scattered across the code base, and in some cases replicated. This code snippet is an example of a 150\% representation, where a product configuration (in terms of macro definitions) excludes some of the syntactic constructs, leaving only those belonging to a specific product.

\begin{lstlisting}[language=Java, caption=Annotative FLDPL, label={lst:Annotative}, float]
class FLD {
	#if defined (LOW_GAS_WARNING)
	const float THRESHOLD;
	#endif

@@

@@

Listing \ref{lst:Delta} shows a Delta-Oriented implementation of FLDPL. The base product here is the Gas FLD. On top of that, deltas for other variants are implemented.

@@

@@

Goal Structuring Notation \cite{Kelly} can be used to layout logical arguments connecting safety goals to pieces of evidence satisfying them. As seen in figure \ref{fig:gsn}, a safety case in GSN is typically composed of 3 kinds of artifacts:

@@

\begin{itemize}
\item Goals: These are the safety goals (requirements) identified as a result of hazard analysis techniques
\item Solutions(Evidences): These are pieces of evidence that a risk/hazard is mitigated. Evidences can be test case results, formal verification outputs, or manual investigation reports
\item Strategies: a strategy provides the logical argument connecting safety goals to their corresponding mitigating evidences
\end{itemize}

%\subsubsection{Structured Assurance Case Metamodel (SACM)}
%
%SACM \cite{SACM} is an Object Management Group (OMG) standard. SACM integrates well with other OMG metamodels (e.g. UML). It provides a much richer notation compared to GSN. The notation is composed of the following metamodels:
%
%\begin{itemize}
%\item An Argumentation Meta-model
%\item An Evidence Meta-model
%\end{itemize}
%
%\section{Adding Variability}
%
%\subsection{GSN Modules and Patterns}
%
%\subsubsection{GSN Modules}
%
%Modules are a GSN extension that facilitate modularity and reuse in safety cases. GSN elements (goals, evidences and strategies) can be packages into modules, and those modules can be reused in different contexts.
%
%\begin{figure}
%  \centering
%  \includegraphics[width=0.75\textwidth]{gsn-modules}
%  \caption{GSN Module Constructs}
%\end{figure}
%
%\subsubsection{GSN Patterns}
%
%GSN patterns are yet another GSN extension. A GSN pattern is a parameterized GSN item. A pattern can then be instantiated by binding a parameter to a concrete value.
%
%\begin{figure}
%  \centering
%  \includegraphics[width=0.75\textwidth]{gsn-patterns}
%  \caption{GSN Pattern Constructs}
%\end{figure}
%
%As an example of a GSN pattern, a generic hazard avoidance pattern would be in the context of an identified hazard, aiming at assuring that the system is safe by addressing that hazard.
%
%\begin{figure}
%  \centering
%  \includegraphics[width=0.75\textwidth]{gsn-pattern-example}
%  \caption{GSN Pattern Example - Hazard mitigation}
%\end{figure}
%
%\subsubsection{GSN Product Line Example}
%
%This is just a fragment of the safety case in \cite{Gallucci_2013}, demonstrating the use of GSN modules:
%
%\begin{figure}
%  \centering
%  \includegraphics[width=0.75\textwidth]{safetyCase}
%  \caption{Safety Case Fragment}
%\end{figure}
%
%%3.2	Comparative Analysis
%%
%%GSN+
%%SACM?
%%GM
%%Expressivity
%%
%%
%%
%%Syntactic Conciseness
%%
%%
%%
%%Well-defined Semantics?
%%
%%
%%
%%Semantic Consistency
%%
%%
%%
%%Traceability within a single product
%%
%%
%%
%%Traceability across products
%%
%%
%%
%%
%

@@

For a specific Fuel Level Display (FLD) product, safety hazards are identified, and then those hazards are structurally addressed in a safety case. For example, a snippet of the FLD safety case addressing Kalman filter hazards is presented in figure \ref{fig:SafetyCaseExample}.

Here the robustness of the estimation algorithm in the context of using the Kalman filter is a goal (G\_6). This goal is refined into a more specific goal: the steadiness of the output of the Kalman filter (G\_7).  This goal is further refined into 3 sub-goals (G\_8, G\_9, G\_10). Goal G\_8 s fulfilled by the results of standard deviation analysis (evidence E\_1). In addition, goals G\_9 and G\_10 are fulfilled by simulation results (evidence E\_2).

 \begin{figure}
  \centering
  \includegraphics[width=0.75\textwidth]{SafetyCaseExample}
  \caption{Kalman filter safety case example}
  \label{fig:SafetyCaseExample}
\end{figure}

@@

\item What is the safety case notation used by GM safety engineers?

\item How is safety case variability currently maintained by GM safety engineers?

\item How are hazards, safety case goals, pieces of evidence managed across different products in a product line by GM safety engineers?

>>>>

Several attempts have been made to leverage the high degree of commonality across products of a Software Product Line (SPL) to minimize the cost of product line analysis in general, and safety analysis in particular. The goals of this document are:

\begin{itemize}
\item Introducing terminology for both Software Product Line and Safety Case concepts
\item Demonstrating the Product Line and Safety Case concepts on a simple exemplar
\item To try to come to a mutual agreement on concepts, terminology and the Product Line safety engineering at GM
\end{itemize}

@@

Throughout this document we use a Fuel Level Display Product Line (FLDPL) example from \cite{Gallucci_2013}. The diagrams in this document also come from \cite{Gallucci_2013} unless otherwise specified.

@@

We start with some basic definitions, together with some simple examples illustrating them.

@@

\item[Software Product Line]
(SPL) is a set of similar software products built from a common set of artifacts (e.g., design/specification models, code constructs, test cases, etc...).

@@

\item[Feature]
is an externally visible property, aspect or quality of a software system.

@@

\item[Feature Diagram]
is a hierarchical graphical representation of a Feature Model, where each non-root feature depends on its parent.

@@

For example, Figure \ref{fig:FeatureDiagram} shows a feature diagram of the FLDPL. In this diagram, the usage context is composed of two mandatory features: Vehicle type and Fuel type. Vehicle type can be either Truck or Bus (mutually exclusive options). Similarly, Fuel type can be either Liquid or Gas. In addition, the FLD system has one mandatory feature (Fuel Level Estimation) and one optional feature (Low Fuel Level Warning). An annotation (O1) is added to the diagram to specify the set of products where the low fuel level warning feature is present.

@@

\item[Presence Condition]
(PC) is a boolean formula over the set of features that represents the set of products in which an artifact exists.
\end{description}

@@

\item[Product Configuration]
is a boolean formula over the set of features of an SPL that specifies the set of features present in a product.
\end{description}

@@

\item[150\% Representation]
is a syntactic representation with syntactic constructs belonging to all the features of the SPL. 
\end{description}

Since in the vast majority of cases all those syntactic constructs cannot all coexist in a product (given feature model constraints), this representation is referred to as 150\%, as opposed to the 100\% representation of a single product.

Examples of 150\% representations can be found in Section 4.

@@

\lstset{
     basicstyle=\ttfamily,
     keywordstyle=\color{blue}\ttfamily,
     stringstyle=\color{red}\ttfamily,
     commentstyle=\color{green}\ttfamily,
     morecomment=[l][\color{magenta}]{\#},
     numbers=left, 
     tabsize=2,
     captionpos=b, 
     frame=shadowbox,
     language=Java,
     morekeywords={if, endif, defined, refines, core, delta, modifies, remove, add, when}
}

@@

\begin{lstlisting}[caption=Generative FLDPL, label={lst:Generative}, float]
// Feature module FLDGas
class FLD {
	void reportFuelLevel(float level) { 
		dashboard.display(level); 
	}
}

@@

\begin{lstlisting}[caption=Generated FLD Product, label={lst:Generated}, float]
class FLD {
	const float THRESHOLD;

@@

Given the product configuration (Liquid $\wedge$ Truck $\wedge$ FuelWarning), a code generator would automatically produce the product in Listing \ref{lst:Generated}.

@@

\begin{lstlisting}[caption=Annotative FLDPL, label={lst:Annotative}, float]
class FLD {
	#if defined (LOW_GAS_WARNING)
	const float THRESHOLD;
	#endif

@@

If we are to rewrite FLDPL using CPP macros, it would look like Listing \ref{lst:Annotative}. A single feature implementation (e.g., LOW\_GAS\_WARNING) is now scattered across the code base (lines 3 and 8). This code snippet is an example of a 150\% representation, where a product configuration (in terms of macro definitions) excludes some of the syntactic constructs, leaving only those belonging to a specific product.

@@

@@

Listing \ref{lst:Delta} shows a Delta-Oriented implementation of FLDPL. The base product here is the Gas FLD (lines 1-7). On top of that, deltas for other variants are implemented. For example, the DLiquidTruck delta (lines 9-15) modifies the FLD class implemented in the base product by removing the reportFuelLevel method and adding another one that passes the fuel level through a Kalman filter. Similarly, DLiquidBus (lines 17-23) and DLowGasWarning (lines 25-35) are deltas for their corresponding features.

@@

Goal Structuring Notation \cite{Kelly} can be used to layout logical arguments connecting safety goals to pieces of evidence satisfying them. As seen in Figure \ref{fig:gsn}, a safety case in GSN is typically composed of 3 kinds of artifacts:

@@

\begin{itemize}
\item Goals: These are the safety goals (requirements) identified as a result of hazard analysis techniques.
\item Solutions (Evidences): These are pieces of evidence that a risk/hazard is mitigated. Evidences can be test case results, formal verification outputs, or manual investigation reports.
\item Strategies: a strategy provides the logical argument connecting safety goals to their corresponding mitigating evidences.
\end{itemize}

%\subsubsection{Structured Assurance Case Metamodel (SACM)}
%
%SACM \cite{SACM} is an Object Management Group (OMG) standard. SACM integrates well with other OMG metamodels (e.g., UML). It provides a much richer notation compared to GSN. The notation is composed of the following metamodels:
%
%\begin{itemize}
%\item An Argumentation Meta-model
%\item An Evidence Meta-model
%\end{itemize}
%
%\section{Adding Variability}
%
%\subsection{GSN Modules and Patterns}
%
%\subsubsection{GSN Modules}
%
%Modules are a GSN extension that facilitate modularity and reuse in safety cases. GSN elements (goals, evidences and strategies) can be packages into modules, and those modules can be reused in different contexts.
%
%\begin{figure}
%  \centering
%  \includegraphics[width=0.75\textwidth]{gsn-modules}
%  \caption{GSN Module Constructs}
%\end{figure}
%
%\subsubsection{GSN Patterns}
%
%GSN patterns are yet another GSN extension. A GSN pattern is a parameterized GSN item. A pattern can then be instantiated by binding a parameter to a concrete value.
%
%\begin{figure}
%  \centering
%  \includegraphics[width=0.75\textwidth]{gsn-patterns}
%  \caption{GSN Pattern Constructs}
%\end{figure}
%
%As an example of a GSN pattern, a generic hazard avoidance pattern would be in the context of an identified hazard, aiming at assuring that the system is safe by addressing that hazard.
%
%\begin{figure}
%  \centering
%  \includegraphics[width=0.75\textwidth]{gsn-pattern-example}
%  \caption{GSN Pattern Example - Hazard mitigation}
%\end{figure}
%
%\subsubsection{GSN Product Line Example}
%
%This is just a fragment of the safety case in \cite{Gallucci_2013}, demonstrating the use of GSN modules:
%
%\begin{figure}
%  \centering
%  \includegraphics[width=0.75\textwidth]{safetyCase}
%  \caption{Safety Case Fragment}
%\end{figure}
%
%%3.2	Comparative Analysis
%%
%%GSN+
%%SACM?
%%GM
%%Expressivity
%%
%%
%%
%%Syntactic Conciseness
%%
%%
%%
%%Well-defined Semantics?
%%
%%
%%
%%Semantic Consistency
%%
%%
%%
%%Traceability within a single product
%%
%%
%%
%%Traceability across products
%%
%%
%%
%%
%

@@

For a specific Fuel Level Display (FLD) product, safety hazards are identified, and then those hazards are structurally addressed in a safety case. For example, a snippet of the FLD safety case addressing Kalman filter hazards is presented in Figure \ref{fig:SafetyCaseExample}.

Here the robustness of the estimation algorithm in the context of using the Kalman filter is a goal (G\_6). This goal is refined into a more specific goal: the steadiness of the output of the Kalman filter (G\_7).  This goal is further refined into 3 sub-goals (G\_8, G\_9, G\_10). Results of standard deviation analysis (evidence E\_1) address Goal G\_8. In addition,  simulation results (evidence E\_2) address goals G\_9 and G\_10.

 \begin{figure}
  \centering
  \includegraphics[width=1.0\textwidth]{SafetyCaseExample}
  \caption{Kalman filter safety case example}
  \label{fig:SafetyCaseExample}
\end{figure}

@@

\item What are the safety case constructs used by GM safety engineers? Is there a standard notation (e.g., GSN) used?

\item How is safety case variability currently managed by GM safety engineers?

\item How are hazards, safety case goals and pieces of evidence managed across different products in a product line by GM safety engineers?

\item How is variability traced throughout the end-to-end Software Lifecycle at GM? In particular, how does safety case variability map to variability in design and code artifacts?
